---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Cassie Lee"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: true
  warning: false

code-fold: true

from: markdown+emoji 
---

::: {.callout-tip icon="false"}
## Github Repo Link

[Cassie's Final Project Github Repo Link](https://github.com/stat301-1-2023-fall/final-project-1-cassielee2025.git)
:::

```{r}
#| label: load-libraries

library(tidyverse)
library(janitor)
library(naniar)
library(skimr)
```

## Data source

This data is from the CDC National Environmental Public Health Tracking Network interactive data explorer.[^1]

[^1]: Link to the interactive data explorer can be found [here](http://www.cdc.gov/ephtracking).

To access the various indicators of air quality, health, and demographics, I had to query and download each each indicator separately. Thus, in my raw data folder, there are 18 different folders representing each query and download. Folders were renamed with a shorthand representing the indicator that is contained in that folder. Each folder has a .csv data file, a .htm footnotes file, and a .pdf general information file.

I will need to join all of the datasets together. Most data files have indicators broken down by each county throughout the nation, however, there are a few indicators that represent data for the entire state. The files that have `county` identifiers also all have a `state` identifier, so data corresponding to states as a whole can be joined using `state`.

## Why this data

I chose this data because I am interested in environmental public health and would like to learn more about air quality and health outcomes. I am currently more familiar with water quality and health outcomes, but I would like to expand my knowledge of environmental exposures through this project.

I originally wanted to look at air quality sensor data throughout Chicago through a project called Array of Things,[^2] but realized that it was time series data. Instead, I shifted my focus from individual sensor data as the unit of measurement to counties across the nation as my unit of observatoin

[^2]: Link to the Array of Things website can be found [here](https://arrayofthings.github.io/index.html).

After looking around several government websites, I realized that a lot of the data that I was interested in regarding various health measurements such as biomarkers and anthropometrics were unavailable to the public. This dataset contains the most relevant information about environmental quality and health outcomes that I was able to access. I selected data from 2018 for most of the indicators, since that was the most complete and most recent data available. There are some indicators that are from 2020.

I believe this data will be sufficiently challenging to work with, as I will have to join 21 different files, there are both numeric and factor variables.

Some interesting questions that I would consider are:

1.  How are various air quality indicators related to each other?

2.  What are the most important indicators for determining the prevalence of asthma, cancer, and chronic obstructive pulmonary disease?

3.  How do demographics like age, gender, race, and ethnicity affect the relationships between air quality and health outcomes?

## Data quality & complexity check

```{r}
#| label: read-in-data

# load in the data
asthma_adult_crude <- read_csv("data/raw/adult-asthma-prevalence-crude/data_135355.csv") %>% 
  clean_names()

asthma_child_crude <- read_csv("data/raw/child-asthma/data_134936.csv") %>% 
  clean_names()

asthma_ed_adjusted <- read_csv("data/raw/asthma-emergency-department-age-adjusted/data_135130.csv") %>% 
  clean_names()

asthma_ed_crude <- read_csv("data/raw/asthma-emergency-department-crude/data_135101.csv") %>% 
  clean_names()

cancer_adjusted <- read_csv("data/raw/cancer-age-adjusted/data_134453.csv") %>% 
  clean_names()

copd_adjusted <- read_csv("data/raw/copd-age-adjusted/data_135253.csv") %>% 
  clean_names()

copd_crude <- read_csv("data/raw/copd-crude/data_135330.csv") %>% 
  clean_names()

days_over_o3_standard <- read_csv("data/raw/days-over-O3-standard/data_135022.csv") %>% 
  clean_names()

days_over_pm_standard <- read_csv("data/raw/days-over-pm-standard/data_134954.csv") %>% 
  clean_names()

age_demographics <- read_csv("data/raw/demographics-age/data_160814.csv") %>% 
  clean_names()

gender_demographics <- read_csv("data/raw/demographics-gender/data_160859.csv") %>% 
  clean_names()

race_ethnicity <- read_csv("data/raw/demographics-race-ethnicity/data_161208.csv") %>% 
  clean_names()

highway_living <- read_csv("data/raw/highway-living/data_151056.csv") %>% 
  clean_names()

highway_schools <- read_csv("data/raw/highway-schools/data_151116.csv") %>% 
  clean_names()

parks_access <- read_csv("data/raw/parks-access/data_143718.csv") %>% 
  clean_names()

pollutants <- read_csv("data/raw/selected-pollutant-concentration/data_134821.csv") %>% 
  clean_names()

socioeconomic_vulnerability <- read_csv("data/raw/socioeconomic-vulnerability-index/data_161326.csv") %>% 
  clean_names()

work_transportation <- read_csv("data/raw/work-transportation/data_151035.csv") %>% 
  clean_names()
```

I am able to read in all 18 data files. In total, I have 150 variables across all the datasets, but each dataset contains several repeated identifiers such as `state_fips`, `state`, `county_fips`, `county`, and `year`.

```{r}
#| label: variable-types
#| results: hide

## age ----
age_demographics %>% 
  skim_without_charts()

## asthma adult crude ----
asthma_adult_crude %>% 
  skim_without_charts()

## asthma child crude ----
asthma_child_crude %>% 
  skim_without_charts() # value is character vector here

# modify value variable in asthma_child_crude
asthma_child_crude %>% 
  arrange(desc(value)) # Suppressed values are changing the value to character

asthma_child_crude <- asthma_child_crude %>% 
  mutate(value = na_if(value, "Suppressed"),
         value = as.numeric(value)) 

asthma_child_crude %>% 
  skim_without_charts()

## asthma ed adjusted ----
asthma_ed_adjusted %>% 
  skim_without_charts() # value is character vector here

# modify value variable in asthma_ed_adjusted
asthma_ed_adjusted %>% 
  arrange(desc(value)) # Suppressed values are changing the value to character

asthma_ed_adjusted <- asthma_ed_adjusted %>% 
  mutate(value = na_if(value, "Suppressed"),
         value = as.numeric(value)) 

asthma_ed_adjusted %>% 
  skim_without_charts()

## asthma ed crude ----
asthma_ed_crude %>% 
  skim_without_charts() # value is character vector

# modify value variable in asthma_ed_crude
asthma_ed_crude %>% 
  arrange(desc(value)) # Suppressed values are changing the value to character

asthma_ed_crude <- asthma_ed_crude %>% 
  mutate(value = na_if(value, "Suppressed"),
         value = as.numeric(value)) 

asthma_ed_crude %>% 
  skim_without_charts()

## cancer adjusted ----
cancer_adjusted %>% 
  skim_without_charts()

## copd adjusted ----
copd_adjusted %>% 
  skim_without_charts()

## copd crude ----
copd_crude %>% 
  skim_without_charts()

## days over o3 standard ----
days_over_o3_standard %>% 
  skim_without_charts()

## days over pm standard ----
days_over_pm_standard %>% 
  skim_without_charts()

## gender demographics ----
gender_demographics %>% 
  skim_without_charts()

## highway living ----
highway_living %>% 
  skim_without_charts()

## highway schools ----
highway_schools %>% 
  skim_without_charts()

## parks access ----
parks_access %>% 
  skim_without_charts()

## pollutants  ----
pollutants %>% 
  skim_without_charts()

## race ethnicity ----
race_ethnicity %>% 
  skim_without_charts()

## socioeconomic vulnerability ----
socioeconomic_vulnerability %>% 
  skim_without_charts()

## work transportation ----
work_transportation %>% 
  skim_without_charts()

```

Each data file has the following variables:

1.  `age_demographics`: 5 categorical, 2 numerical, and 1 logical

2.  `asthma_adult_crude`: 5 categorical, 4 numerical, and 2 logical

3.  `asthma_child_crude`: 2 categorical, 2 numerical, and 2 logical

4.  `asthma_ed_adjusted`: 5 categorical, 2 numerical, and 1 logical

5.  `asthma_ed_crude`: 5 categorical, 2 numerical, and 1 logical

6.  `cancer_adjusted`: 2 categorical, 2 numerical, and 2 logical

7.  `copd_adjusted`: 5 categorical, 4 numerical, and 2 logical

8.  `copd_crude`: 5 categorical, 4 numerical, and 2 logical

9.  `days_over_o3_standard`: 4 categorical, 2 numerical, and 2 logical

10. `days_over_pm_standard`: 4 categorical, 2 numerical, and 2 logical

11. `gender_demographics`: 5 categorical, 2 numerical, and 1 logical

12. `highway_living`: 4 categorical, 2 numerical, and 2 logical

13. `highway_schools`: 4 categorical, 2 numerical, and 2 logical

14. `parks_access`: 5 categorical, 2 numerical, and 1 logical

15. `pollutants`: 5 categorical, 2 numerical, and 1 logical

16. `race_ethnicity`: 5 categorical, 2 numerical, and 1 logical

17. `socioeconomic_vulnerability`: 4 categorical, 2 numerical, and 2 logical

18. `work_transportation`: 5 categorical, 3 numerical, and 1 logical

There are also several variables throughout the datasets such as `data_comment`, `x8`, and `x11` that are completely missing. These variables are classified as logical variables in `skimr`. `asthma_ed_crude` `asthma_ed_adjusted`, and `asthma_ed_crude` have "Suppressed" values that are missing. Additionally, some of the indicators I selected were only available at the state level, so every county in each state will have the same measure for those indicators.

## Potential data issues

For the final project, I intend for each county in the US to be an observation. However, the demographic data I downloaded considers each specific demographic in each county as an observation. For example, in `age_demographics`, each `county` plus `age_group` is an observation with a measured value. For each demographic variable will likely have to create some kind of proxy variable that indicates the dominant demographic group within that variable for each county. I may also be able to examine the distribution of each demographic variable and create a cutoff for a relatively "high" or "low" proportion of the population with in each demographic group.

Additionally, in `race_ethnicity`, the US census definitions of race and ethnicity are combined under one variable, so there is a single category for each race plus ethnicity. I will need to use string transformations to separate the two.

Finally, several indicators are only reported by state, so every county in each state will have the same value for the given indicator. This greatly limits the variability in the data, and may affect bi and multivariate analysis.

## Miscellaneous 

**Proposed timeline:**

11/4/23: univariate analysis and determining how I will create and use proxy variables for demographics

11/11/23: merge datasets and half of bivariate analysis

11/18/23: second half of bivariate analysis and half of multivariate analysis with various demographics as the third variable

11/25/23: second half of multivariate analysis and writing

11/29/23: writing and turn in