---
title: "Air Quality Indicators and Disease Prevalence Across the US"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Cassie Lee"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

code-fold: true

from: markdown+emoji 
---

::: {.callout-tip icon="false"}
## Github Repo Link

[Cassie's Final Project Github Repo Link](https://github.com/stat301-1-2023-fall/final-project-1-cassielee2025.git)
:::

```{r}
#| label: load-pckgs-data

library(tidyverse)
library(skimr)
library(sf)
library(patchwork)
load("data/full_air_quality_data.rda")
```

# Introduction

Air pollution is the presence of sufficient quantities of contaminants in the atmosphere for a duration that is long enough to cause harm to human health.[^1] Air pollution mainly enters the body through the lungs, and while it mostly impacts the heart, lungs, and brain, it has the potential to affect other organs as well by traveling through the bloodstream.[^2]

This EDA explores the relationships between air quality indicators, certain lung diseases, and how sociodemographic vulnerabilities affect the relationships between air quality and health in the United States (excluding territories). The goal of this analysis is to build a deeper understanding of how various air pollutants impact health across the United States.

The data was queried at the county level and downloaded from the CDC National Environmental Public Health Tracking Network interactive data explorer.[^3] This network was built to centralize environmental health data on the national, state, and county level across the United States.[^4]

The three central questions explored in this analysis were:

1.  How are various air and environmental quality indicators related to each other?

2.  What are the most important indicators for determining the prevalence of asthma, cancer, and chronic obstructive pulmonary disease?

3.  How do sociodemographics like age, gender, race, and socioeconomic vulnerability affect the relationships between air quality and health outcomes?

[^1]: [Air quality and health](https://www.who.int/teams/environment-climate-change-and-health/air-quality-and-health/health-impacts)

[^2]: [Air quality and health](https://www.who.int/teams/environment-climate-change-and-health/air-quality-and-health/health-impacts)

[^3]: [CDC National Environmental Public Health Tracking Network](http://www.cdc.gov/ephtracking)

[^4]: [CDC National Environmental Public Health Tracking Network](http://www.cdc.gov/ephtracking)

# Data overview & quality

The air pollutants selected includes the days over the ozone standard, the percent of days over the PM 2.5 standard, benzene, formaldehyde, acetaldehyde, carbon tetrachloride, and 1,3-butadiene pollution. Environmental quality indicators selected includes the percent of people living near highways, the percent of public schools near highways, access to parks, and methods of transportation to work (walking, biking, driving alone, carpooling, public transportation, and none).

The indicators selected for prevalence of lung diseases includes the crude prevalence of adult and child asthma, the crude and age adjusted rates of emergency department visits for asthma, age adjusted rates of lung and bronchus cancer, and the crude and age adjusted rates of chronic pulmonary obstructive disease. 

The indicators of sociodemographic data selected includes age, gender, race, and and the social vulnerability index[^5]. Sociodemographic characteristics of each county was compared to other counties in the United States and used to identify counties with comparatively higher percentages of vulnerable groups (Appendix A). The majority race demographic in each county was also identified.

If available, all data was downloaded at the county level for 2018. Crude rates of child asthma and age adjusted rates of lung and bronchus cancer were only available at the state level. Data for the usual method of transportation to work was downloaded for the time period of 2017 to 2021. 

The final merged dataset includes 42 variables and 3144 observations matching each county and Washington DC. There are 4 identifying variables, 5 factor variables, 32 numeric variables, and one simple features geometry variable for mapping. 

With the exception of Rio Arriba, Mexico, all counties have complete sociodemographic information. Rio Arriba, Mexico is missing data for the social vulnerability index. The prevalence of adult asthma is complete across all observations, but 1402 observations are missing for the prevalence of child asthma and 1789 observations are missing for crude and age adjusted emergency department visits for asthma. The majority of counties do not have information about the days over the ozone and PM 2.5 air quality standards. 1567 observations are missing data for the percent of public school near highways. All other variables are either complete or missing at most 1 observation.

[^5]: [CDC/ATSDR Social Vulnerability Index](https://www.atsdr.cdc.gov/placeandhealth/svi/index.html)

# Explorations

## Indicators of air and environmental quality

As seen in @days-over-distribution, the distribution of days over the ozone and particulate matter size 2.5 microns (PM 2.5) standards are both skewed right by observations with significantly more days over the standard. For both ozone and PM 2.5, most of the counties experienced no more than 20 days over the standard. However, the degree of right skewing for the distribtion of days over ozone standard is significantly higher than for the distribution of days over the PM 2.5 standard. 

@days-over-relationship shows the relationship between the days over the ozone and PM 2.5 standard.

```{r}
#| label: days-over-distribution
#| fig-cap: Distribution of days over air quality standards.

p1 <- full_data %>% 
  ggplot(aes(days_over_o3_standard)) +
  geom_histogram(binwidth = 10, boundary = 0, fill = "#f56042") +
  theme_minimal() +
  labs(
    title = "Days over ozone standard",
    x = "Days",
    y = NULL
  )

p2 <- full_data %>% 
  mutate(pm_standard = days_over_pm_standard/100 * 365) %>% 
  ggplot(aes(pm_standard)) +
  geom_histogram(binwidth = 10, boundary = 0, fill = "#f5a442") +
  theme_minimal() +
  labs(
    title = "Days over PM 2.5 standard",
    x = "Days",
    y = NULL
  )

p1 + p2
```

```{r}
#| label: days-over-relationship
#| fig-cap: Relationship between days over PM 2.5 standard and days over ozone standards excluding outliers (over 50 days over one or both standards)

full_data %>% 
  mutate(pm_standard = days_over_pm_standard/100 * 365) %>% 
  ggplot(aes(days_over_o3_standard, pm_standard)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  lims(
    x = c(0, 50),
    y = c(0, 50)
  ) +
  theme_minimal() +
  labs(
    title = "Days over PM 2.5 standard vs days over ozone standard",
    x = "Days ozone standard",
    y = "Days over PM 2.5 standard"
  )
```




# Appendix
## A. Identifying sociodemographic vulnerabilities
